Radiocarbon measurement of CO~2~ from carbonate minerals using a hybrid gas ion source with an open split interface

### 2022-04-12 doc/AMS-15/AMS-15_paper/AMS_HGIS_els/AMS_HGIS_els.Rmd
Add MK edits.

---
title: "Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface"
author:
  - name: Brett Longworth
    email: blongworth@whoi.edu
    affiliation: Woods Hole Oceanographic Institution
    correspondingauthor: true
  - name: Josh Burton
    email: jburton@whoi.edu
    affiliation: Woods Hole Oceanographic Institution
  - name: Simon Pendleton
    email: simon.pendleton@plymouth.edu
    affiliation:  Woods Hole Oceanographic Institution,Plymouth State University
  - name: Sydney Moser
    email: smose007@uottawa.ca
    affiliation: University of Ottawa
  - name: Mark Roberts
    email: mroberts@whoi.edu
    affiliation: Woods Hole Oceanographic Institution
  - name: Mark Kurz
    email: mkurz@whoi.edu
    affiliation: Woods Hole Oceanographic Institution
address:
  - code: Woods Hole Oceanographic Institution
    address: Department of Geology and Geophysics, 360 Woods Hole Rd., Woods Hole, MA, 02540
  - code: Plymouth State University
    address: Environmental Science and Policy, 17 High St., Plymouth, MA, 03264
  - code: University of Ottawa
    address: Geography, Environment and Geomatics, 550 Cumberland St., Ottawa, ON, Canada, K1N 6N5
abstract: |
  We have developed a method for measuring radiocarbon in carbonate minerals as CO~2~ gas via a NEC MCSNICS hybrid sputter gas ion source (HGIS). The method uses helium as a carrier gas to displace CO~2~ from sample vials to an open split, where a glass capillary samples the mixture for delivery directly to the HGIS. This method skips the gas transfer and quantification steps used in a closed inlet HGIS system, simplifying sample measurement. Samples larger than 8 mg carbonate can be measured. Results from measurements of consensus standards (TIRI-I, IAEA C2, and an internal modern shell standard), and samples from a marine core (F^14^C = 0.4 - 1.15) show that the method agrees well with traditional AMS measurement of the same samples as graphite, and that the 1$\sigma$ uncertainty is about 1 %. We discuss advantages and disadvantages of continuous flow sample introduction, and the effect of reduced precision on calibrated age-depth models produced using gas-source data.
keywords: 
  - radiocarbon
  - carbonate
  - hybrid gas ion source
journal: "Nuclear Instruments and Methods in Physics Research Section B"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: ../HGIS_paper.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
csl: elsevier-with-titles.csl
output: 
  rticles::elsevier_article:
    latex_engine: xelatex
    keep_tex: true
    citation_package: default
---

@@

Radiocarbon Accelerator Mass Spectrometry (AMS) is a sensitive method for determining the age, provenance, and composition of carbon-containing materials. A primary disadvantage of the method is its cost, which often limits the number of samples analyzed. This forces researchers to carefully choose which samples to analyze and limits repeat measurements for error determination. One driver of the cost of AMS analysis is the time and effort required for sample preparation, including reduction of the sample to graphite. Analysis of CO~2~ directly via a gas-accepting AMS ion source eliminates the cost and effort of graphitization and returns results more rapidly. The NOSAMS USAMS system at the National Oceanographic AMS lab (NOSAMS) is based on a 3MV Tandetron accelerator [@Longworth2015] combined with a National Electrostatics Corporation (NEC) 134-sample gas-capable MCSNICS source [@Xu2007] . We present here a system for semi-automated analysis of carbonate minerals by acid hydrolysis in vials and introduction to the ion source in a flow of helium via an open split.

Most interfaces to gas-accepting sputter ion sources are a closed inlet design [@Xu2007; @Uhl2007; @ruff_gas_2007; @ruff_-line_2010; @Fahrni2013; @wacker_versatile_2013; @gottschalk_radiocarbon_2018]. A mixture of sample and helium is introduced to the source through a capillary, and flow is controlled by the pressure in the inlet volume and the length and diameter of the capillary. Pressure is adjusted by changing either the inlet volume or the amount of gas in the inlet. We designed a open inlet where a capillary samples gas for delivery to the ion source in a flowing stream that exhausts to the atmosphere, often called an open split. This defines the delivery pressure as 1 ATM. CO~2~ flow to the source can be controlled by changing the length or diameter of the capillary, or by changing the ratio of CO~2~ to He in the sample stream.

@@

The gas-capable 134-sample MCSNICS from National Electrostatics Corporation (NEC) [@Xu2007] works on the same principle as hybrid sources developed by ETH Zürich [@ruff_gas_2007; @Ruff2010] and High Voltage Engineering Europa (HVEE) [@BronkRamsey2004]. The sample CO~2~ - helium mixture is introduced onto a titanium insert or frit in an aluminum target cathode using stainless tubing mounted to an arm with a probe that mates with the back of the cathode. Sample gas is exposed to a beam of charged cesium atoms, which causes dissociation of CO~2~ and ionization of molecular and atomic species. Negative ions are extracted from the target, carbon species are selected, and interfering masses filtered out by the USAMS Tandetron beamline.

@@

The open split interface (Figure \ref{fig1}) supplies a helium - CO~2~ mixture to a split where a capillary delivers gas to the source and the rest is vented. Flowing helium is used to displace sample CO~2~ from the sample vial to the delivery capillary in the split. Our open-split interface is constructed of simple, off-the-shelf parts. The core of the design is a dual needle and split housed in a Valco tee fitting. Displacement helium flows in through the sidearm of the tee and into the sample vial through 1/16 in. stainless steel tubing, which forms the outer flow path of the concentric dual needle.

The mixture of sample CO~2~ and displacement helium from the vial flows up a 21 gauge (0.819 mm) needle inserted through the tee and stainless tubing which forms the inner flow path of the dual needle. The open split is a VICI Luer adapter (part ZBUMLPK) with 5cm of 1/16 in. PFTE tubing attached. The split is connected to the top of the inner needle, and the source delivery capillary is inserted into the split. The split has fairly low volume (< 1 mL). The system was operated manually during development, but may be mounted on the arm of a vial-sampling autosampler to enable automated operation.

@@

Reference and process blank materials were not pretreated. Carbonate shell samples for the core comparison study were picked from a split core and cleaned by washing with distilled water and dried at 60 C. Shell samples were crushed and mixed, but not otherwise pretreated.

@@

Samples were measured on a modified 3MV Tandetron AMS system [@Longworth2015]. Measurement of a sample proceeded as follows: a clean sputter target cathode was rotated into position and the target was pre-cleaned by sputtering for 5 minutes or until cathode ^12^C^-^ current fell below 100 nA. The delivery needle was then transferred to the sample vial, which started gas delivery to the split and source. Ion current was allowed to stabilize for 30 s, at which point the data acquisition was started. Data was acquired in 10 blocks of 90 s each, for 15 min total acquisition time. The displacement helium was allowed to flow constantly during cathode cleaning and analysis, and the sample needle was inserted into a helium-filled vial between samples. Normalizing standards, blanks, and secondary standards were measured identically to unknowns. Total time for a measurement was 20-25 min including target pre-cleaning and changing samples.

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO~2~ connected through a regulator and glass capillary to the ion source to measure source performance over a range of CO~2~ flows. Best ^12^C^-^ ion current for the source was 10 $\mu$A, or roughly 10% of graphite target current. Peak current was produced at a CO~2~ flow rate of 7.5 $\mu$l min^-1^ CO~2~, but the source produced near optimal currents with flows from 7.5-25 $\mu$l min^-1^ CO~2~. This is comparable with other hybrid sources [@Xu2007], although improved target design and continued optimization of source parameters may improve performance [@Salazar2013; @Fahrni2013]. Typical currents for this study were between 4-9 $\mu$A and average current for all samples was `r curmean` ± `r cursd` $\mu$A. Several studies have shown best currents at lower CO~2~ flow rates [@Xu2007; @Ruff2010].

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="80%", fig.cap = "\\label{fig2}Performance of the NOSAMS MCSNICS hybrid gas ion source. (a) Measured $^{12}C^-$ ion current during displacement of a pure $CO_2$ vial sample with helium. (b) Modeled flows of helium and $CO_2$ to the ion source during the same dilution experiment. (c) Measured ion current for modeled $CO_2$ flow. The stability of the source over a range of flow produces stable current for up to 40 min of measurement time.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig2_current_flow.pdf"))
```

Source ion currents are variable by day and by target under similar source, inlet, and sample conditions. Factors affecting source current may include capillary condition, source and gas delivery arm alignment, slight variability in target geometry, source condition, and performance of the ionizer and cesium. Capillary clogging caused occasional flow issues, especially when sampling from vials containing acid.

@@

Where $C(t)$ is the fraction CO~2~, $r$ is the rate of helium displacement flow, and $V$ is the initial volume of the CO~2~-filled vial. Figure 2c shows the resulting relationship of sample ion current to CO~2~ flow rate. The most important feature of the current-flow relationship is the wide range of CO~2~ flow rates over which stable ion current is measured, which in turn produces relatively stable currents and ratios despite a changing carbon flow rate during measurement. Several studies have measured peak currents over a narrower range of flows [@Fahrni2013; @broek_optimization_nodate] at high helium dilution rates (> 10:1), while Xu et. al [@Xu2007] shows a somewhat broader peak with pure CO~2~. Broek et. al [@broek_optimization_nodate] found that the range of flow rates with peak currents increased with decreasing dilution ratio. While lower rates of dilution used in our HGIS method may decrease current slightly, the observed peak broadening allows stable measurement over the range of CO~2~ flows produced by the open split.

Work with currents and flows (above and Figure 2) shows that stable currents are produced until less than 2 ml CO~2~ (0.8 mg C) remains in the sample vial. Tests of a range of sample sizes support this estimate and show that no reduction in measured current is seen in samples larger than 1 mg C which corresponds to 8 mg CaCO~3~ sample material. To ensure ample sample for optimal measurements, we chose 15 mg as a target sample mass. To ensure that samples started at or slightly above atmospheric pressure, we injected helium with a syringe to equilibrate pressure in the vials when the reaction with acid was complete.

@@

The HGIS method produced higher ratios for radiocarbon dead samples than the graphite method. The average uncorrected $F^{14}C$ for radiocarbon-free reference materials IAEA C1 marble and TIRI F spar calcite was `r blankmean` ± `r blanksd` (1$\sigma$ uncertainty, N=`r nblank`). We investigated the sources of this method blank as a current-dependent blank and as a current-independent blank, and implemented a blank correction for the method.

@@

```{r fig3, fig.width = 5, fig.height = 5, fig.align='center', out.width="80%", fig.cap = "\\label{fig3}Assessment of blanks. (a) Current dependent blank model and data. $F^{14}C$ and current of measurements from dilution tests of modern and dead tank standards are used to generate linear fits which should cross at the $F^{14}C$ and current of the constant current blank. Direct determination of $F^{14}C$ and current of the blank using clean targets and helium without CO~2~ are also shown. (b) Fraction modern of large IAEA C1 and TIRI-F carbonate blank samples compared to sample current. Horizontal and vertical bars represent measurement uncertainty in current and ratio, respectively.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig3_blanks.pdf"))
```

@@

While dilution and direct measurement data show evidence for a constant blank with a variable effect depending on sample size, the blank correction may be simplified if blanks, standards, and unknown samples are the same size [@donahue_isotope-ratio_1990; @roberts_14c_2019]. If the fraction modern of the blank is 0.8 modern as estimated above, radiocarbon-free samples will show the most effect of a constant contamination blank with current. Figure 3b shows the fraction modern of measured blanks compared to their mean ^12^C^-^ ion current. Fraction modern of measured blanks does decrease with increasing current, but the decrease is small, less than F^14^C 0.01 over the range of currents measured. This change with current is roughly equal to the variability between blanks, so the error due to the use of a current-dependent blank based on samples measured over multiple wheels will likely be larger than a current independent blank correction based on blanks measured on each wheel. Constant current blank is likely the best model, but we simplify to a linear correction based on the measured ratio of process blanks and normalizing standards:

@@

Reference materials analyzed with the hybrid gas method were in good agreement with consensus values. Mean difference from consensus for all measurements was F^14^C `r format(cons_sum$fm_diff_mean, digits = 1, nsmall = 4)` shows that the hybrid method has no bias in aggregate. The standard deviation of the difference in F^14^C was `r format(cons_sum$fm_diff_sd, nsmall = 4)`, which may be taken as an estimate of the 1$\sigma$ precision of the method. The mean of propagated measurement errors for these samples was `r format(cons_sum$sig_fm_corr_mean, nsmall = 4)` F^14^C for these samples, which is in good agreement with the method precision above. The method error is about three times larger than the roughly 3 ‰ precision of graphite measurements, but is within the range of method precision of other gas sources.

@@

```{r fig5, fig.width = 5, fig.height = 5, fig.align='center', out.width="80%", fig.cap = "\\label{fig5}Comparison of Haiti core samples measured as graphite and via HGIS. (a) Difference between HGIS and graphite measurements of the same materials. Difference between HGIS and graphite measurements is shown on the vertical axis and plotted against the mean of the two measurements. Samples are offset on the horizontal axis for clarity. Error bars are 1 $\\sigma$ measurement uncertainty for each sample. Graphite measurements are included to allow comparison of uncertainty. The bottom row compares age-depth models produced using HGIS (b) and graphite measurements (c). Grey subplots show probability distribution of calibrated ages.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig5_core_carbonates.pdf"))
```

@@

Age models were developed for both the HGIS measurements (Figure 5b) and the traditional graphite measurements (Figure 5c) of the core carbonates. We used the Bchron Baesian radiochronology package [@haslett_simple_2008] to develop these models. We applied a correction for the regional marine reservoir effect with a $\Delta$ R of -216 ± 92 years based on the Oxcal model [@ramsey_bayesian_2009] with 10 reservoir points. Samples with pre-bomb radiocarbon ages were calibrated using the Marine20 curve @heaton_marine20marine_2020. The NH zone 2 curve from Hua et. al @hua_atmospheric_2013 was used for calibration of postmodern dates, but these measurements were not included in the age model due to uncertainties in reservoir correction. The age-depth curves produced by the two datasets are very similar, with the largest differences being in the uncertainty of the individual dates.

@@

We developed a method for the analysis of carbonates using an open-split interface to a hybrid sputter gas ion source. This system produces results that agree with graphite measurements within the approximately 1% method error from 8-20 mg of carbonate. Sample preparation takes about 10 min. of person time and 6 h of automated processing and reaction time, and 20 min. of measurement time per sample. The simplicity of of this approach reduces sample cost by eliminating graphitization, and may improve reliability and reduce maintenance compared to more complex sample inlet designs.

@@

This work was supported by the NOSAMS NSF cooperative agreement 1755125. We are grateful to K. von Reden, A. Gagnon, K. Gospodinova, and J. Hlavenka for initial work and discussions on the NOSAMS HGIS; C. McIntyre, T. Broek, G. Salazar, and T. Ognibene for discussions on gas handling and gas ion sourcery; J. Gunnard for improvements to the text; and the staff of NOSAMS for assistance with the HGIS and preparation of graphite samples.

>>>>

---
title: "Radiocarbon measurement of CO~2~ from carbonate minerals using a hybrid gas ion source with an open split interface"
author:
  - name: Brett Longworth
    email: blongworth@whoi.edu
    affiliation: Woods Hole Oceanographic Institution
    correspondingauthor: true
  - name: Josh Burton
    email: jburton@whoi.edu
    affiliation: Woods Hole Oceanographic Institution
  - name: Simon Pendleton
    email: simon.pendleton@plymouth.edu
    affiliation:  Woods Hole Oceanographic Institution,Plymouth State University
  - name: Sydney Moser
    email: smose007@uottawa.ca
    affiliation: University of Ottawa
  - name: Mark Roberts
    email: mroberts@whoi.edu
    affiliation: Woods Hole Oceanographic Institution
  - name: Mark Kurz
    email: mkurz@whoi.edu
    affiliation: Woods Hole Oceanographic Institution
address:
  - code: Woods Hole Oceanographic Institution
    address: Department of Geology and Geophysics, 360 Woods Hole Rd., Woods Hole, MA, 02540
  - code: Plymouth State University
    address: Environmental Science and Policy, 17 High St., Plymouth, MA, 03264
  - code: University of Ottawa
    address: Geography, Environment and Geomatics, 550 Cumberland St., Ottawa, ON, Canada, K1N 6N5
abstract: |
  We have developed a method for measuring radiocarbon in carbonate minerals as CO~2~ gas via a NEC MCSNICS hybrid sputter gas ion source (HGIS). The method uses helium as a carrier gas to displace CO~2~ from sample vials to an open split, where a glass capillary samples the mixture for delivery directly to the HGIS. This method skips the gas transfer and quantification steps used in a closed inlet HGIS system, simplifying sample measurement. Samples larger than 8 mg carbonate can be measured. Results from measurements of consensus standards (TIRI-I, IAEA C2, and an internal modern shell standard), and samples from a marine core (F^14^C = 0.4 - 1.15) show that the method agrees well with traditional AMS measurement of the same samples as graphite, and that the 1$\sigma$ uncertainty is about 1 %. We discuss advantages and disadvantages of continuous flow sample introduction, and the effect of reduced precision on calibrated age-depth models produced using gas-source data.
keywords: 
  - radiocarbon
  - carbonate
  - hybrid gas ion source
journal: "Nuclear Instruments and Methods in Physics Research Section B"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: ../HGIS_paper.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
csl: elsevier-with-titles.csl
output: 
  rticles::elsevier_article:
    latex_engine: xelatex
    keep_tex: true
    citation_package: default
---

@@

Radiocarbon Accelerator Mass Spectrometry (AMS) is a sensitive method for determining the age, provenance, and composition of carbon-containing materials. A primary disadvantage of the method is cost, which can limit the number of samples analyzed. This forces researchers to carefully select samples for analysis and may limit repeat measurements for error determination. One driver of the cost of AMS analysis is the time and effort required for sample preparation, including reduction of the sample to graphite. Analysis of CO~2~ directly via a gas-accepting AMS ion source eliminates the cost and effort of graphitization and returns results more rapidly. The NOSAMS USAMS system at the National Oceanographic AMS lab (NOSAMS) is based on a 3MV Tandetron accelerator [@Longworth2015] combined with a National Electrostatics Corporation (NEC) 134-sample gas-capable MCSNICS source [@Xu2007] . We present here a system for semi-automated analysis of carbonate minerals by acid hydrolysis in vials and introduction to the ion source in a flow of helium via an open split.

Most interfaces to gas-accepting sputter ion sources are a closed inlet design [@Xu2007; @Uhl2007; @ruff_gas_2007; @ruff_-line_2010; @Fahrni2013; @wacker_versatile_2013; @gottschalk_radiocarbon_2018]. A mixture of sample and helium is introduced to the source through a capillary, and flow is controlled by the pressure in the inlet volume and the length and diameter of the capillary. Pressure is adjusted by changing either the inlet volume or the amount of gas in the inlet. We designed a open inlet where a capillary samples gas for delivery to the ion source in a flowing stream that exhausts to the atmosphere, referred to here as an open split. This defines the delivery pressure as 1 ATM. CO~2~ flow to the source can be controlled by changing the length or diameter of the capillary, or by changing the ratio of CO~2~ to He in the sample stream.

@@

The gas-capable 134-sample MCSNICS from National Electrostatics Corporation (NEC) [@Xu2007] works on the same principle as hybrid sources developed by ETH Zürich [@ruff_gas_2007; @Ruff2010] and High Voltage Engineering Europa (HVEE) [@BronkRamsey2004]. The sample CO~2~ - helium mixture is introduced onto a titanium insert or frit in an aluminum target cathode using stainless tubing mounted to an arm with a probe that mates with the back of the cathode. Sample gas is exposed to a beam of charged cesium atoms as it flows over the titanium insert, which causes dissociation of CO~2~ and ionization of molecular and atomic species. Negative ions are extracted from the target, carbon species are selected, and interfering masses filtered out by the USAMS Tandetron beamline.

@@

The open split interface (Figure \ref{fig1}) supplies a helium - CO~2~ mixture to a split where a capillary transfers gas to the ion source and the gas not taken by the capillary is vented. Flowing helium is used to displace sample CO~2~ from the sample vial to the delivery capillary in the split. The open-split interface is constructed of simple, off-the-shelf parts. The core of the design is a dual needle and split housed in a Valco tee fitting. Displacement helium flows in through the sidearm of the tee and into the sample vial through 1/16 in. stainless steel tubing, which forms the outer flow path of the concentric dual needle.

The mixture of sample CO~2~ and displacement helium from the vial flows up a 21 gauge (0.819 mm) needle inserted through the tee and stainless tubing which forms the inner flow path of the dual needle. The open split is a VICI Luer adapter (part ZBUMLPK) with 5cm of 1/16 in. PFTE tubing attached. The split is connected to the top of the inner needle, and the source delivery capillary is inserted into the split. The split has fairly low volume (< 1 mL). The system was operated manually during development, but could be mounted on the arm of a vial-sampling autosampler to enable automated operation.

@@

Reference and process blank materials were not pretreated. Carbonate shell samples for the core comparison study were picked from a split core and cleaned by washing with distilled water and dried at 60 C. Shell samples were crushed and mixed, but not otherwise pretreated. The homogenized carbonates were sub-sampled for measurement by HGIS (15mg) and as graphite (10mg).

@@

The HGIS measurements were carried out on a modified 3MV Tandetron AMS system [@Longworth2015]. Measurement of a sample proceeded as follows: a clean sputter target cathode was rotated into position and the target was pre-cleaned by sputtering for 5 minutes or until cathode ^12^C^-^ current fell below 100 nA. The delivery needle was then transferred to the sample vial, which started gas delivery to the split and source. Ion current was allowed to stabilize for 30 s, at which point the data acquisition was started. Data was acquired in 10 blocks of 90 s each, for 15 min total acquisition time. The displacement helium was allowed to flow constantly during cathode cleaning and analysis, and the sample needle was inserted into a helium-filled vial between samples. Normalizing standards, blanks, and secondary standards were measured identically to unknowns. Total time for a measurement was 20-25 min including target pre-cleaning and changing samples. Graphite samples were prepared at NOSAMS and measured on the CFAMS system using standard procedures [@gagnon_nosams_2000; @roberts_high-performance_2010].

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO~2~ connected through a regulator and glass capillary to the ion source to measure source performance over a range of CO~2~ flows. Best ^12^C^-^ ion current for the source was 10 $\mu$A, or roughly 10% of typical graphite target current. Peak current was produced at a CO~2~ flow rate of 7.5 $\mu$l min^-1^ CO~2~, but the source produced near optimal currents with flows from 7.5-25 $\mu$l min^-1^ CO~2~. This is comparable with other hybrid sources [@Xu2007], although improved target design and continued optimization of source parameters may improve performance [@Salazar2013; @Fahrni2013; @broek_optimization_nodate]. Typical currents for this study were between 4-9 $\mu$A and average current for all samples was `r curmean` ± `r cursd` $\mu$A. Several studies have shown best currents at lower CO~2~ flow rates [@Xu2007; @Ruff2010].

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="80%", fig.cap = "\\label{fig2}Performance of the NOSAMS MCSNICS hybrid gas ion source. (a) Measured $^{12}C^-$ ion current during displacement of modern tank $CO_2$ from a sample vial with helium. (b) Modeled flows of helium and $CO_2$ to the ion source during the same dilution experiment. (c) Measured ion current for the modeled $CO_2$ flow from b. The stability of the source over a range of flow produces stable current for up to 40 min of measurement time.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig2_current_flow.pdf"))
```

Source ion currents are variable by day and by target under similar source, inlet, and sample conditions. Factors affecting source current may include capillary condition, source and gas delivery arm alignment, slight variability in target geometry, source condition, and performance of the ionizer and cesium reservoir. Capillary clogging caused occasional flow issues, especially when sampling from vials containing acid.

@@

Where $C(t)$ is the fraction CO~2~, $r$ is the rate of helium displacement flow, and $V$ is the initial volume of the CO~2~-filled vial. Figure 2c shows the resulting relationship of sample ion current to CO~2~ flow rate. The most important feature of the current-flow relationship is the wide range of CO~2~ flow rates over which stable ion current is measured, which in turn produces relatively stable currents and ratios despite a changing carbon flow rate during measurement. Several studies have measured peak currents over a narrower range of flows [@Fahrni2013; @broek_optimization_nodate] at high helium dilution rates (> 90% helium), while Xu et. al [@Xu2007] shows a somewhat broader peak with pure CO~2~. Broek et. al [@broek_optimization_nodate] found that the range of flow rates with peak currents increased with decreasing dilution ratio. While lower rates of dilution used in our HGIS method may decrease current slightly, the observed relationship between ion current and flow rate allows stable measurement over the range of CO~2~ flows produced by the open split.

Work with currents and flows (above and Figure 2) shows that stable currents are produced until less than 2 ml CO~2~ (0.8 mg C) remains in the sample vial. Tests of a range of sample sizes support this estimate and show that no reduction in measured current is seen in samples larger than 1 mg C which corresponds to 8 mg CaCO~3~ sample material. To ensure sufficient sample for optimal measurements, we chose 15 mg as a target sample mass. To ensure that samples started at or slightly above atmospheric pressure, we injected helium with a syringe to equilibrate pressure in the vials when the reaction with acid was complete.

@@

The HGIS method produced higher $^{14/12}C$ ratios for radiocarbon dead samples than the graphite method. The average uncorrected $F^{14}C$ for radiocarbon-free reference materials IAEA C1 marble and TIRI F spar calcite was `r blankmean` ± `r blanksd` (1$\sigma$ uncertainty, N=`r nblank`), compared to XXX for graphite. We investigated the sources of this method blank as a current-dependent blank and as a current-independent blank, and implemented a blank correction for the method.

@@

```{r fig3, fig.width = 5, fig.height = 5, fig.align='center', out.width="80%", fig.cap = "\\label{fig3}Assessment of blanks. (a) Current dependent blank model and data. $F^{14}C$ and current of measurements from dilution tests of modern and dead tank standards are used to generate linear fits which should cross at the $F^{14}C$ and current of the constant current blank. Direct determination of $F^{14}C$ and current of the blank using clean targets and helium without $CO_2$ are also shown. (b) Fraction modern of large IAEA C1 and TIRI-F carbonate blank samples compared to sample current. Horizontal and vertical bars represent measurement uncertainty in current and ratio, respectively.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig3_blanks.pdf"))
```

@@

While dilution and direct measurement data show evidence for a constant blank with a variable effect depending on sample size, the blank correction may be simplified if blanks, standards, and unknown samples are the same size [@donahue_isotope-ratio_1990; @roberts_14c_2019]. If the fraction modern of the blank is 0.8 modern as estimated above, radiocarbon-free samples will show the most effect of a constant contamination blank with current. Figure 3b shows the fraction modern of measured blanks compared to their mean ^12^C^-^ ion current. Fraction modern of measured blanks does decrease with increasing current, but the decrease is small, less than F^14^C 0.01 over the range of currents measured. This change with current is roughly equal to the variability between blanks, so the error due to the use of a current-dependent blank based on samples measured over multiple measurement sessions will likely be larger than a current independent blank correction based on blanks measured during each session. Constant current blank is likely the best model, but we simplify to a linear correction based on the measured ratio of process blanks and normalizing standards:

@@

Reference materials analyzed with the hybrid gas method were in good agreement with consensus values. Mean difference from consensus for all measurements was F^14^C `r format(cons_sum$fm_diff_mean, digits = 1, nsmall = 4)` shows that the hybrid method has no bias in aggregate. The standard deviation of the difference in F^14^C was `r format(cons_sum$fm_diff_sd, nsmall = 4)`, which may be taken as an estimate of the 1$\sigma$ precision of the method. The mean of propagated measurement errors for these samples was `r format(cons_sum$sig_fm_corr_mean, nsmall = 4)` F^14^C for these samples, which is in good agreement with the method precision above. The method error is about three times larger than the 2--3 ‰ precision of graphite measurements, but is within the range of method precision of other gas sources.

@@

```{r fig5, fig.width = 5, fig.height = 5, fig.align='center', out.width="80%", fig.cap = "\\label{fig5}Comparison of Haiti core samples measured as graphite and via HGIS. (a) Difference between HGIS and graphite measurements of the same materials. Difference between HGIS and graphite measurements is shown on the vertical axis and plotted against the mean of the two measurements. Error bars are 1 $\\sigma$ measurement uncertainty for each sample. Graphite measurements are included to allow comparison of uncertainty. Age-depth models produced using HGIS (b) and graphite measurements (c). Grey subplots show probability distribution of calibrated ages. (d) Combined age-depth prediction and 95\\% confidence intervals for HGIS and graphte. (e) Offset of HGIS and graphite values from the mean of the models with confidence intervals.", echo = FALSE}
knitr::include_graphics(here("doc/AMS-15/AMS-15_paper/figures/fig5_core_carbonates.pdf"))
```

@@

Age models were developed for both the HGIS measurements (Figure 5b) and the traditional graphite measurements (Figure 5c) of the core carbonates. We used the Bchron Baesian radiochronology package [@haslett_simple_2008] to develop these models. We applied a correction for the regional marine reservoir effect with a $\Delta$ R of -216 ± 92 years based on the Oxcal model [@ramsey_bayesian_2009] with 10 reservoir points. Samples with pre-bomb radiocarbon ages were calibrated using the Marine20 curve @heaton_marine20marine_2020. The NH zone 2 curve from Hua et. al @hua_atmospheric_2013 was used for calibration of postmodern dates, but these measurements were not included in the age model due to uncertainties in reservoir correction. The age-depth curves produced by the two datasets agree within the uncertainty of the chronologies, with the largest differences being in the uncertainty of the individual dates.

@@

We developed a method for the analysis of carbonates using an open-split interface to a hybrid sputter gas ion source. This system produces results that agree with graphite measurements within the approximately 1% method error from 8-20 mg of carbonate. Sample preparation takes about 10 min. of person time and 6 h of automated processing and reaction time, and 20 min. of measurement time per sample. The simplicity of of this approach can potentially reduce sample cost by eliminating graphitization, and may improve reliability and reduce maintenance compared to more complex sample inlet designs.

@@

This work was supported by the NOSAMS NSF cooperative agreement 1755125. S. Pendleton was supported by a WHOI/NOSAMS Postdoctoral fellowship. We are grateful to K. von Reden, A. Gagnon, K. Gospodinova, and J. Hlavenka for initial work and discussions on the NOSAMS HGIS; C. McIntyre, T. Broek, G. Salazar, and T. Ognibene for discussions on gas handling and gas ion sourcery; J. Gunnard and T. Broek for improvements to the text; and the staff of NOSAMS for assistance with the HGIS and preparation of graphite samples.
